A Generic and Efficient E-field Parallel Imaging Correlator for Next-Generation Radio Telescopes

### 2016-08-14 imaging_demo/ms.tex
Make minor revisions throughout
From 3f838e8 (mnras):
====================

In ms.tex,

(*) Make minor revisions throughout

In epic.bib,

(*) Add reference to HERA paper by DeBoer et al. (2016)

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregularly arranged heterogeneous antenna. We have demonstrated its versatility to image with heterogeneous arrays by showing that it robustly estimates the input sky model while wrong assumptions about array homogeneity lead to severe mis-estimates. EPIC is highly modular and parallelizable, implemented in object oriented Python, and publicly available. We have verified the images produced to be equivalent to those produced using traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying EPIC. Inherent availability of calibrated time-domain images on digitizer writeout time-scales and vastly lower I/O bandwidth relative to visibility-based systems will make it a prime candidate for transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to
thousands of individual antennas are needed to achieve desired survey speeds.
Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study
of the history of hydrogen gas throughout the universe's evolution is pushing
technology development towards arrays of low-cost antennas with large fields of
view and densely packed layouts. Similarly, the search for transient objects
and regular monitoring of the time-dependent sky is driving instruments in the
same direction with the added requirement of fast read-outs. A number of new 
telescopes are being or were developed around the world based on this new 
paradigm, including the Hydrogen Epoch of Reionization 
Array\footnote{http://reionization.org} (HERA), the Murchison Widefield Array 
(MWA;\citealt{tin13,bow13}), the Precision Array for Probing the Epoch of 
Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray 
(LOFAR;\citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment 
(CHIME,\citealt{ban14}), the Long Wavelength Array (LWA, \citealt{ell13}), and 
the low frequency Square Kilometer Array (SKA1-Low \citealt{mel13}).

@@

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log_2 \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical beam patterns, and is provably 
optimal \citep{mor11}. This algorithm uses the antenna beam patterns to 
grid the electric field measurements to a regular grid in the software 
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from traditional visibility-based techniques.

@@

We provide a brief summary of the mathematical equivalence of the MOFF and 
FX correlators detailed in \citet{mor11}. We first relate the dirty image produced 
from visibilities to the electric fields of astrophysical sources, then show that 
operations can be reordered to produce the same images at a lower 
computational cost.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. {\bf The noise-like cross-terms ideally have zero mean}. This expression can be further simplified with the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)= \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{W}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power response, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by the multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  V_{ab}(f,t) &= \int W_{ab}(\mathbf{r}_{ab}^\prime-\mathbf{r},f,t) \nonumber\\ 
              &\qquad \times \left[\int I(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}.\hat{\mathbf{s}}/c}\,\dif\Omega\right]\dif^2\mathbf{r}\,+\, n_{ab}(f,t),
\end{align}
where, $W_{ab}(\mathbf{r},f,t)$ denotes the {\bf power response of the antenna pair obtained in the measurement plane} by a spatial Fourier transform of $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$. Effectively, the multiplication in image space by $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ has been replaced by a convolution with $W_{ab}(\mathbf{r},f,t)$ in {\bf measurement plane}. This is the software holographic equivalent of a traditional FX correlator output.

@@

We use the EPIC simulator to generate stochastic electric field samples from a 
sky model consisting of 10 point sources of flux densities 10~Jy each at random 
locations. In our simulations, we use 64 frequency channels each of width 
$\Delta f = 40$~kHz. The number of time stamps integrated in one integration 
cycle was kept at eight where each A/D timeseries is $1/\Delta f=25\,\mu$s long. 
We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 
tiles within a square bounding box of 150~m on each side were used. We assumed 
all tiles are identical and have a square shaped electric field illumination 
footprint 4.4~m on each side. Besides the stochastic sky noise present in the 
simulated electric fields, no noise from the instrument is added.

@@

Before the outputs can be compared, we describe the elimination of a minor 
mathematical difference between the two techniques. The squaring operation under 
MOFF imaging in the image plane introduces antenna auto-correlations around the 
zero spacing in the $uv$-plane which are absent in traditional visibility-based 
imaging. In order to facilitate a robust comparison between MOFF and 
visibility-based imaging techniques, these auto-correlations are removed from 
the MOFF algorithm output, which is otherwise not an essential part of the core 
algorithm. We describe below how they are removed.

@@

We inverse Fourier transform the squared images and beams back to the $uv$ 
plane and subtract the estimated auto-correlation kernel scaled to the peak 
value centred at the zero spacing pixel. The final averaged image is obtained 
by Fourier transforming the $uv$ plane data and weights with the 
auto-correlations subtracted to the image plane. These images are now comparable 
to those obtained from visibility-based imaging. This step of removing 
auto-correlations is optional and required to be performed only once per 
integration time-scale and does not add significant cost to the full operation.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the dirty images (top) and synthesized beams
(bottom) obtained with antenna-based MOFF and FX visibility-based imaging 
algorithms packaged in EPIC. The antenna auto-correlations that correspond to 
zero spacing have been removed from the correlated weights and data in the 
$uv$ plane, the MOFF image and the corresponding synthesized beam as 
described in \S\ref{sec:rm-autocorr}. The sky positions of the simulated sources 
are indicated by solid black circles. The reconstructed sky image has the 
simulated sources at the expected sky positions in either case. Both algorithms 
result in images and synthesized beams that are well matched with each other. 
Their fluxes are modulated by a multiplicative power pattern corresponding to 
that of a uniform square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure5}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Dirty images (top) and synthesized beams (bottom) obtained from 
    simulated data using EPIC implementation of antenna-based MOFF algorithm 
    (left) and visibility-based imaging (right). The solid black circles in the
    top panels indicate the simulated source positions. The antenna 
    auto-correlations at zero-spacing have been removed from the MOFF images. 
    The images in either case reconstruct the sources at the right locations 
    with the fluxes expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

\begin{figure}
  \includegraphics[width=\columnwidth]{figure6}
  \caption{.}
  \label{fig:f-engine}
\end{figure}

Now we investigate the gridded cross-correlation weights in the $uv$-plane. 
Fig.~\ref{fig:MOFF-FX-uvwts} shows the correlated weights obtained with MOFF 
imaging (left) and visibility-based imaging (right). There are two notable 
differences. The first is in the weights around zero spacing. Though both show a
dominant void around zero-spacing, the void obtained with MOFF algorithm shows 
many pixels with non-zero weights. In contrast, the zero-spacing void from 
traditional imaging consists of predominantly zero-valued pixels. The second 
difference is in the weights outside the zero-spacing void. In some regions they 
are different from each other at the few percent level though the sum of weights 
in these regions is the same. For instance, note the difference in weights at 
$(u,v)\approx (2.5,4)$. The reason for these differences is described below.

@@

\begin{figure}
  \includegraphics[width=\columnwidth]{figure7}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.eps}
  \caption{.}
  \label{fig:MOFF-FX-uvwts}
\end{figure}

We study the effect of the differences in gridded weights on the image plane. 
Fig.~\ref{fig:psf-diff} shows the difference between the synthesized beams 
obtained with the two methods. A difference map between the two synthesized 
beams is shown on top. The amplitude of the difference appears to be 
modulated by the directional power response of the antenna. At the bottom, 
in radial bins, the rms of the synthesized beam (gray) and the rms of the 
difference map (black) are plotted in percentage units relative to the peak 
(to be read using the axis on the left side of the plot). The antenna power 
pattern (red; to be read using the scale on the right) is plotted for reference.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure8}
  % \includegraphics[width=\columnwidth]{diff_psf_MOFF-FX_test_aperture}
  % \includegraphics[width=\columnwidth]{diff_psf_MOFF-FX_test_aperture.eps}
  \caption{Map of difference between the synthesized beams obtained with the 
    two methods (left) and radial statistics of the synthesized beams and 
    their differences (right). The radial variations of rms of the synthesized 
    beam (gray) and that of the rms of the difference (black) are shown as 
    percentage of the peak synthesized beam. Radially averaged directional 
    antenna power response in absolute scale is shown in red and is to be read 
    with the scale on the right side of the axis. The maximum difference is of 
    the order of a few percent. Amplitude of the rms of the difference is 
    modulated by the power pattern of the antenna.}
  \label{fig:psf-diff}
\end{figure}

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power pattern. 
The rms of differenced synthesized beams is also modulated by the antenna power 
pattern. The rms of the difference is definitely lesser than the rms of the 
synthesized beam in the central regions up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This 
implies that the beams are well matched in the central regions. In the outer 
regions, their mismatch is comparable to the rms of synthesized beams. This 
indicates the two synthesized beams are not completely randomly different from 
each other in which case the rms of the difference would have been 
$\approx \sqrt{2}$ higher than the rms of the each of the synthesized beams. 
This indicates that while differences exist, large fractions of them are still 
well matched to each other even out to the horizon. Thus the rounding errors in 
gridding do not affect the statistics of the images or the synthesized beams.

@@

We corrected the cable delays, but otherwise assume the data is sufficiently 
calibrated to image directly. A test of EPICal on this data will be conducted 
in the future.

@@

We present a methodology to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array and compare it with that from standard interferometry that assumes a homogeneous array (all antennas are identical). We also demonstrate how the assumption in the latter results in a mis-estimation of recovered flux densities of sources.

>>>>

% Abstract of the paper
\begin{abstract}
Modern radio telescopes are favouring densely packed array layouts with large numbers of antennas ($\Nant\gtrsim 1000$). Since the complexity of traditional correlators scales as $\mathcal{O}(\Nant^2)$, there will be a steep cost for realizing the full imaging potential of these powerful instruments. Through our generic and efficient E-field Parallel Imaging Correlator (EPIC), we present the first software demonstration of a generalized direct imaging algorithm, namely, the Modular Optimal Frequency Fourier (MOFF) imager. It takes advantage of the multiplication-convolution theorem of Fourier transforms. Not only does it bring down the cost for dense layouts to $\mathcal{O}(\Nant\log_2\Nant)$ but can also image from irregular layouts and heterogeneous arrays of antennas. We demonstrate its versatility to image with heterogeneous arrays by showing that it robustly estimates the input sky model while wrong assumptions about array homogeneity lead to severe mis-estimates. EPIC is highly modular and parallelizable, implemented in object oriented Python, and publicly available. We have verified the images produced to be equivalent to those produced using traditional techniques to within a precision set by gridding coarseness. We have also validated our implementation on data observed with the Long Wavelength Array (LWA). Antenna layouts with dense filling factors consisting of a large number of antennas such as LWA, the Square Kilometre Array, Hydrogen Epoch of Reionization Array, and Canadian Hydrogen Intensity Mapping Experiment will gain significant computational advantage by deploying EPIC. Inherent availability of calibrated time-domain images on digitizer writeout time-scales and vastly lower I/O bandwidth relative to visibility-based systems will make it a prime candidate for transient searches of Fast Radio Bursts (FRB) as well as planetary and exoplanetary phenomena. 
\end{abstract}

@@

Radio astronomy is entering an era in which interferometers of hundreds to thousands of individual antennas are needed to achieve desired survey speeds. Nowhere is this more apparent than at radio frequencies below 1.4 GHz. The study of the history of hydrogen gas throughout the universe's evolution is pushing technology development towards arrays of low-cost antennas with large fields of view and densely packed layouts. Similarly, the search for transient objects and regular monitoring of the time-dependent sky is driving instruments in the same direction with the added requirement of fast read-outs. A number of new telescopes are being or were developed around the world based on this new paradigm, including the Hydrogen Epoch of Reionization Array\footnote{http://reionization.org} (HERA; \citealt{deb16}), the Murchison Widefield Array (MWA; \citealt{tin13,bow13}), the Precision Array for Probing the Epoch of Reionization (PAPER; \citealt{par10}), the LOw Frequency ARray (LOFAR; \citealt{van13}), the Canadian Hydrogen Intensity Mapping Experiment (CHIME; \citealt{ban14}), the Long Wavelength Array (LWA; \citealt{ell13}), and the low frequency Square Kilometer Array (SKA1-Low; \citealt{mel13}).

@@

Direct imaging correlators have begun to be explored on deployed arrays including
the Basic Element for SKA Training II (BEST-2) array \citep{fos14}, the Omniscope
\citep{zhe14}, and an earlier incarnation at higher frequencies with the intent
of pulsar timing \citep{oto94, dai00}. However, each of these examples make
assumptions about the redundancy of the array layout, and require the collecting
elements are identical. On the other hand, the MOFF algorithm achieves the same
$\Ngrid \log_2 \Ngrid$ computational scaling without placing any restriction on
antenna placement, can accommodate non-identical antennas, and is provably 
optimal \citep{mor11}. This algorithm uses the antenna beam patterns to 
grid the electric field measurements to a regular grid in the software 
holography/A-transpose fashion \citep{mor09,bha08,teg97b} before performing the
spatial FFT. This process has been shown to theoretically produce a data product
identical to images produced from traditional visibility-based techniques.

@@

We provide a brief summary of the mathematical equivalence of the MOFF and 
FX correlators detailed in \citet{mor11}. We first relate the image produced 
from visibilities to the electric fields of astrophysical sources, then show that 
operations can be reordered to produce the same images at a lower 
computational cost.

@@

Interferometers measure {\it visibilities} -- the degree of coherence between electric fields measured by a pair of antennas \citep{van34,zer38,tho01}. A visibility, $V_{ab}$, can be written as:
\begin{align}
  V_{ab}(f,t) &= \left\langle E_a(f,t)E_b^\star(f,t) \right\rangle_t \label{eqn:cc-vis}\hfill\\
              &= \left\langle \left[ \int \mathcal{W}_a(\hat{\mathbf{s}},f,t)\,\mathcal{E}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega\right.\right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left. n_a(f,t) \vphantom{\int}\right] \nonumber\\
              &\quad \times \left[ \int \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t)\,\mathcal{E}^\star (\hat{\mathbf{s'}},f,t)\,e^{2\pi i f\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}}/c}\,\dif\Omega^\prime \right. \nonumber\\
              &\qquad\qquad\qquad\qquad\qquad\qquad\qquad\quad + \left.\left. n^\star_b(f,t) \vphantom{\int}\right] \right\rangle_t \nonumber\\
              &= \iint \mathcal{W}_a(\hat{\mathbf{s}},f,t) \mathcal{W}^\star_b(\hat{\mathbf{s'}},f,t) \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t) \mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \nonumber\\
              &\qquad\qquad\qquad\qquad \times e^{-2\pi i f(\mathbf{r}_a\!\cdot\,\hat{\mathbf{s}}-\mathbf{r}_b\!\cdot\,\hat{\mathbf{s'}})/c}\,\dif\Omega\,\dif\Omega^\prime \nonumber\\
              &\qquad +\textrm{noise-like cross-terms.}
\end{align}
where we have brought the time average into the integral under the assumption that the aperture illumination pattern does not change over the time-scale of the averaging and $\star$ denotes complex conjugation. The noise-like cross-terms ideally have zero mean. This expression can be further simplified with the sky brightness, $\mathcal{I}(\hat{\mathbf{s}},f,t)= \left\langle \mathcal{E}(\hat{\mathbf{s}},f,t)\,\mathcal{E}^\star(\hat{\mathbf{s'}},f,t) \right\rangle_t \delta(\hat{\mathbf{s}}-\hat{\mathbf{s'}})$, and defining the antenna-pair sky power response function (or the directional antenna-pair power pattern), $\mathcal{W}_{ab}(\hat{\mathbf{s}})~\equiv~\mathcal{W}_a(\hat{\mathbf{s}},f,t)~\mathcal{W}^\star_b(\hat{\mathbf{s}},f,t)$. The result is the visibility expressed in terms of the sky brightness, the power pattern, and uncorrelated noise terms which we group into $n_{ab}(f,t)$.
\begin{align}\label{eqn:measurement-eqn-1}
  V_{ab}(f,t) &= \int \mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)\,\mathcal{I}(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega \nonumber\\
  &\qquad\qquad + n_{ab}(f,t),
\end{align}
where, the baseline coordinate $\mathbf{r}_{ab}\equiv\mathbf{r}_a-\mathbf{r}_b$ is the vector separation between the two antennas. This signifies that the visibility ($V_{ab}$) measured between a pair of antennas is obtained by the multiplying the sky brightness $\mathcal{I}(\hat{\mathbf{s}},f,t)$ by the antenna power response $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ and Fourier transforming from the directional coordinates ($\hat{\mathbf{s}}$) to {\bf the measurement plane}, which are then sampled at the locations of the antenna spacings (or baselines), namely, $\mathbf{r}_{ab}$, and added to the noise $n_{ab}$.

This can be equivalently re-written as:
\begin{align}\label{eqn:software-holography}
  V_{ab}(f,t) &= \int W_{ab}(\mathbf{r}_{ab}-\mathbf{r},f,t) \nonumber\\ 
              &\qquad \times \left[\int I(\hat{\mathbf{s}},f,t)\,e^{-2\pi i f\mathbf{r}.\hat{\mathbf{s}}/c}\,\dif\Omega\right]\dif^2\mathbf{r}\,+\, n_{ab}(f,t),
\end{align}
where, $W_{ab}(\mathbf{r},f,t)$ denotes the {\bf power response of the antenna pair obtained in the measurement plane} by a spatial Fourier transform of $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$. Effectively, the multiplication in image space by $\mathcal{W}_{ab}(\hat{\mathbf{s}},f,t)$ has been replaced by a convolution with $W_{ab}(\mathbf{r},f,t)$ in {\bf measurement plane}. This is the software holographic equivalent of a traditional FX correlator output.

@@

We use the EPIC simulator to generate stochastic electric field samples from a sky model consisting of 10 point sources of random flux densities $>10$~Jy each at random locations. In our simulations, we use 64 frequency channels each of width $\Delta f = 40$~kHz. The number of time stamps integrated in one integration cycle was kept at eight where each A/D timeseries is $1/\Delta f=25\,\mu$s long. We use the MWA array layout \citep{bea12} for demonstration. Only the inner 51 tiles within a square bounding box of 150~m on each side were used. We assumed all tiles are identical and have a square shaped electric field illumination footprint 4.4~m on each side. Besides the stochastic sky noise present in the simulated electric fields, no noise from the instrument is added.

@@

Before the outputs can be compared, we describe the elimination of a distinct difference between the two techniques. The squaring operation under MOFF imaging in the image plane introduces antenna auto-correlations around the zero spacing in the $uv$-plane which are absent in traditional visibility-based imaging. In order to facilitate a robust comparison between MOFF and visibility-based imaging techniques, these auto-correlations are removed from the MOFF algorithm output, which is otherwise not an essential part of the core algorithm. We describe below how they are removed.

@@

We inverse Fourier transform the squared images and beams back to the $uv$ plane and subtract the estimated auto-correlation kernel scaled to the peak value centred at the zero spacing pixel. The final averaged image is obtained by Fourier transforming the $uv$ plane data and weights with the auto-correlations subtracted to the image plane. These images are now comparable to those obtained from visibility-based imaging. This step of removing auto-correlations needs to be performed only once per integration time-scale and does not add significant cost to the full operation.

@@

Fig.~\ref{fig:MOFF-FX-image} shows the weighted dirty images (top) and synthesized beams (bottom) obtained with antenna-based MOFF and FX visibility-based imaging algorithms packaged in EPIC. The antenna auto-correlations that correspond to zero spacing have been removed from the correlated weights and data in the $uv$ plane, the MOFF image and the corresponding synthesized beam as described in \S\ref{sec:rm-autocorr}. The reconstructed sky image has the simulated sources at the expected sky positions in either case. Both algorithms result in images and synthesized beams that are well matched with each other. As expected, their fluxes are attenuated by a weighting proportional to the square of the antenna power pattern corresponding to a uniform square aperture.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure5}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_10_random_source_positions_4_iterations_test_aperture_zoomed.eps}
  \caption{Weighted dirty images (top) and synthesized beams (bottom) obtained from simulated data using EPIC implementation of antenna-based MOFF algorithm (left) and visibility-based imaging (right). The antenna auto-correlations at zero-spacing have been removed from the MOFF images. The images in either case reconstruct the sources at the right locations with the fluxes attenuated expected after multiplication by the antenna power pattern. 
    The synthesized beams from the two algorithms are well matched in size and
    shape. The overall modulation by the power pattern is seen clearly in both
    images.}
  \label{fig:MOFF-FX-image}
\end{figure}

We examine in detail the respective synthesized beams in each case in Fig.~\ref{fig:MOFF-FX-psf-diff}. Slices at $m=0$ of the synthesized beams weighted by the antenna power pattern are shown for MOFF method using EPIC (solid black) and visibility-based (dashed gray) imaging. The two are found to match well. A magnified view shows that some differences at the level of $<0.5$\% exist in some regions. These are attributed to differences in the $uv$-plane antenna cross-correlation weights in the two methods which in turn arises due to the amount of coarseness in grid spacing as described below.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure6}
  \caption{Synthesized beams weighted by antenna power pattern sliced at $m=0$ obtained with MOFF algorithm using EPIC (solid black line) and visibility-based imaging (dashed gray line). The two appear almost identical. The inset provides a magnified view of differences in the synthesized beam slices between the two techniques at levels $\lesssim 0.5$\% relative to the peak. These are attributed to differences that arise in gridding and depends on the coarseness of the grid.}
  \label{fig:MOFF-FX-psf-diff}
\end{figure}

The left panel of Fig.~\ref{fig:MOFF-FX-uvwts-diff} shows differences (in percentage relative to peak) in cross-correlated weights obtained with MOFF imaging in EPIC and visibility-based imaging. The maximum difference appears near the center of the $uv$-plane corresponding to antenna auto-correlations which have not been perfectly removed in the former. In other regions of the $uv$-plane, the differences are of the order of a few percent. The reason for these differences is described below.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure7}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed}
  % \includegraphics[width=\columnwidth]{MOFF_FX_comparison_uvwts_test_aperture_zero_spacing_removed.eps}
  \caption{{\it Left}: Differences (in \%) in cross-correlated weights in the $uv$ plane relative to the peak. The biggest difference ($\sim 20$\%) is found around the zero-spacing corresponding to auto-correlations of antenna weights due to gridding differences augmented by the relatively high number density of antenna auto-correlation footprints in that region. In most of the other regions, differences of the order of less than a few percent are seen. {\it Right}: Histogram (expressed as percentage) of the relative differences (in \%) between antenna cross-correlation weights shown on the left binned in intervals of 0.5\%. Over 70\% of the $uv$-cells do not differ by more than 0.5\% and over 90\% of $uv$-cells only differ by $<5$\% for the gridding coarseness used.} 
  \label{fig:MOFF-FX-uvwts-diff}
\end{figure}

@@

We study the effect of the differences in gridded weights on the image plane. Fig.~\ref{fig:image-psf-diff} shows the difference between the synthesized beams obtained with the two methods. A difference map between the two synthesized beams is shown on the left. The amplitude of the difference appears to be modulated by the directional power response of the antenna. On the right, in radial bins, the rms of the synthesized beam (gray) and the rms of the difference map (black) are plotted in percentage units relative to the peak (to be read using the axis on the left side of the plot). The antenna power pattern (red; to be read using the scale on the right) is plotted for reference.

\begin{figure}
  \includegraphics[width=\columnwidth]{figure8}
  % \includegraphics[width=\columnwidth]{diff_psf_MOFF-FX_test_aperture}
  % \includegraphics[width=\columnwidth]{diff_psf_MOFF-FX_test_aperture.eps}
  \caption{Map of difference between the synthesized beams obtained with the 
    two methods (left) and radial statistics of the synthesized beams and 
    their differences (right). The radial variations of rms of the synthesized 
    beam (gray) and that of the rms of the difference (black) are shown as 
    percentage of the peak synthesized beam. Radially averaged directional 
    antenna power response in absolute scale is shown in red and is to be read 
    with the scale on the right side of the axis. The maximum difference is of 
    the order of a few percent. Amplitude of the rms of the difference is 
    modulated by the power pattern of the antenna.}
  \label{fig:image-psf-diff}
\end{figure}

The synthesized beam rms is proportional to the antenna power pattern as 
expected from a point spread function uncorrected for the antenna power pattern. 
The rms of differenced synthesized beams is also modulated by the antenna power 
pattern. The rms of the difference is definitely lesser than the rms of the 
synthesized beam in the central regions up to $(l^2+m^2)^{1/2}\lesssim 0.3$. This 
implies that the beams are well matched in the central regions. In the outer 
regions, their mismatch is comparable to the rms of synthesized beams. These findings are consistent with Fig.~\ref{fig:MOFF-FX-psf-diff}. This 
indicates the two synthesized beams are not randomly different from 
each other in which case the rms of the difference would have been 
$\approx \sqrt{2}$ higher than the rms of the each of the synthesized beams. 
This means that while differences exist, large fractions of them are still 
well matched to each other even out to the horizon. Thus the rounding errors in 
gridding do not affect the statistics of the images or the synthesized beams.

@@

We corrected the cable delays, but otherwise assume the data is sufficiently calibrated to image directly. A detailed demonstration of EPICal on this data is presented in \citet{bea16}.

@@

We present a methodology to understand the effective angular weighting in the image obtained with EPIC from a heterogeneous array and compare it with that from standard interferometry that assumes a homogeneous array (all antennas are identical). We also demonstrate how the assumption in the latter results in a mis-estimation of the recovered sky model.

Dropping the noise term and time dependence, the visibility measured by an antenna pair can be written from equation~\ref{eqn:measurement-eqn-1} as:
\begin{align}\label{eqn:measurement-eqn-2}
  V_{ab}(f) &= \int \mathcal{W}^\textrm{I}_a(\hat{\mathbf{s}},f)\mathcal{W}^{\textrm{I}\star}_b(\hat{\mathbf{s}},f)\,\mathcal{I}(\hat{\mathbf{s}},f)\,e^{-2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}\,\dif\Omega,
\end{align}
where, superscript `$\textrm{I}$' in the $\mathcal{W}$ term indicates the weighting is inherently introduced by the instrument during the measurement process. The beam-weighted dirty image is:
\begin{align}
  \mathcal{I}_\textrm{D}(\hat{\mathbf{s}},f) &= \frac{1}{\Nant(\Nant-1)/2}\,\sum_{ab}\, V_{ab}(f)\,e^{2\pi i f\mathbf{r}_{ab}\!\cdot\,\hat{\mathbf{s}}/c}.
\end{align}
